---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![DOI](https://zenodo.org/badge/doi/10.5281/zenodo.48423.svg)](http://dx.doi.org/10.5281/zenodo.48423)

| Branch  | Build status | Code coverage |
| ------- | ------------ | ------------- |
| Master  | [![wercker status](https://app.wercker.com/status/5f154cd480de20b68cde62705e817436/s/master "wercker status")](https://app.wercker.com/project/bykey/5f154cd480de20b68cde62705e817436) | [![codecov.io](https://codecov.io/github/inbo/multimput/coverage.svg?branch=master)](https://codecov.io/github/inbo/multimput?branch=master) |
| Develop | [![wercker status](https://app.wercker.com/status/5f154cd480de20b68cde62705e817436/s/master "wercker status")](https://app.wercker.com/project/bykey/5f154cd480de20b68cde62705e817436) | [![codecov.io](https://codecov.io/github/inbo/multimput/coverage.svg?branch=develop)](https://codecov.io/github/inbo/multimput?branch=develop) |

CAUTION: GitHub flavoured markdown doesn't support the rendering of mathematics at this moment. Hence the mathematics in this README are not rendered properly. The information below is also available as a vignette within the package. The mathematics will be rendered in the vignette. To read the vignette one needs to install the package first. 

# Installation instructions

This package requires the `INLA` package. You need to install it with `install.packages("INLA", repos = "https://www.math.ntnu.no/inla/R/stable")`. If this fails you can use `devtools::install_github("INBO-BMK/INLA")`. Note that the latter is just a read-only mirror which is infrequently updated. Hence installing `INLA` from https://www.math.ntnu.no/inla is highly recommended.

When `INLA` is installed, we can install `multimput` with `devtools::install_github("INBO-BMK/multimput", build_vignettes = TRUE)`. To view the vignette use `vignette("Impute", package = "multimput")`

A docker image with all the required dependencies is available from https://hub.docker.com/r/inbobmk/multimput/. Use `docker pull inbobmk/multimput` to get it.

```{r child="vignettes/child/details.Rmd"}
```
