---
title: "Clamping imputations"
author: "Thierry Onkelinx"
date: "2026-10-05"
bibliography: multimput.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clamping imputations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{ggplot2}
---

```{r}
#| label: settings
#| echo: false

library(knitr)
opts_chunk$set(message = FALSE, fig.width = 7.1, fig.height = 5)
```

# Clamping

`multimput` supports clamping as of version `0.3.0`.
Clamping allows to set lower and or upper boundaries for both the imputed values as for their aggregations.

# Clamping at the observation level

A minimal example is a site that was partly inaccessible and therefore we possibly missed some of the animals.
Thus the observed count is rather a minimum than a correct estimate.
In this case we set the observation to missing, impute the observation and clamp the imputations to at least equal to the observed minimum.

```{r}
#| label: prepare-data

set.seed(20261005)
n_year <- 10
n_site <- 10
prob_missing <- 0.1
dataset <- expand.grid(year = seq_len(n_year), site = seq_len(n_site))
dataset$id <- seq_along(dataset$year)
rf_site <- rnorm(n_site)
dataset$eta <- 2 + log(1.1) * dataset$year + rf_site[dataset$site]
dataset$y <- rpois(nrow(dataset), lambda = exp(dataset$eta))
dataset$value <- ifelse(
  rbinom(nrow(dataset), size = 1, prob = prob_missing),
  dataset$y,
  NA
)
dataset$y[!is.na(dataset$value)] <- NA
dataset$eta <- NULL
```

Prepare the imputation model and generate the imputations with and without clamping.
Note that the clamped imputations are, as expected, larger that the unclamped imputations.

```{r}
#| label: raw-impute

library(lme4)
library(multimput)
model <- glmer(y ~ year + (1 | site), data = dataset, family = poisson)
no_minimum <- impute(model = model, dataset)
minimum <- impute(model = model, dataset, minimum = "value")
data.frame(
  unclamped = rowMeans(no_minimum@Imputation),
  clamped = rowMeans(minimum@Imputation)
) |>
  round(2)
data.frame(
  unclamped = apply(no_minimum@Imputation, 1, min),
  clamped = apply(minimum@Imputation, 1, min)
) |>
  round(2)
```

# Clamping at the aggregation level

A more complex example has a design with locations and sublocations.
The count should be collected at the sublocations level but some of the historical data is only available at the location level.
Furthermore we know that we have no guarantee that every sublocation of a location is surveyed during every visit.
When we have only a total at the location level, we don't have information about which sublocations were surveyed or not.

Let's create a dataset with 10 years, 20 locations and 5 sublocations per location.

```{r}
#| label: prepare-data-2

n_year <- 10
n_location <- 20
n_sublocation <- 5
rf_location <- rnorm(n_location)
rf_sublocation <- rnorm(n_location * n_sublocation)

dataset <- expand.grid(
  location = seq_len(n_location),
  sublocation = seq_len(n_sublocation),
  year = seq_len(n_year)
)
dataset$sublocation <- interaction(dataset$location, dataset$sublocation)
dataset$id <- seq_along(dataset$year)
dataset$eta <- 2 +
  log(1.1) * dataset$year +
  rf_location[dataset$location] +
  rf_sublocation[dataset$sublocation]
dataset$y <- rpois(nrow(dataset), lambda = exp(dataset$eta))
dataset$eta <- NULL
dataset$location <- as.character(dataset$location)
dataset$sublocation <- as.character(dataset$sublocation)
```

Now let's poke some holes in the dataset by randomly selecting 10% of the year-location combinations.
For the selected combinations we will set the counts of all sublocations to missing.

```{r}
#| label: clamps

prob_missing <- 0.1
only_totals <- sample(
  n_year * n_location,
  size = floor(n_year * n_location * prob_missing)
)
minimum <- aggregate(y ~ year + location, data = dataset, sum)
minimum <- minimum[
  sample(nrow(minimum), size = floor(nrow(minimum) * prob_missing)),
]
dataset <- merge(
  dataset,
  minimum,
  by = c("year", "location"),
  all.x = TRUE,
  suffixes = c("", ".min")
)
dataset[!is.na(dataset$y.min), "y"] <- NA
dataset$y.min <- NULL
```

Next we impute the missing values and aggregate the imputations at the location level.
Then we do the same but we clamp the aggregated imputations to the minimum value for each year-location combination.
This figure shows the estimated effect of year on the log(count) for both the clamped and unclamped imputations.
Note that the clamped imputations are higher than the unclamped imputations, because we are enforcing a minimum value.

```{r}
#| label: fig-impute-aggregate
#| fig-cap: Estimated year effects with and without clamping.

model <- glmer(
  y ~ year + (1 | location) + (1 | sublocation),
  data = dataset,
  family = poisson
)
imputed <- impute(model = model, dataset)
unclamped <- aggregate_impute(
  object = imputed,
  grouping = c("year", "location"),
  fun = sum
)
clamped <- aggregate_impute(
  object = imputed,
  grouping = c("year", "location"),
  fun = sum,
  minimum = minimum
)
extractor <- function(model) {
  summary(model)$coefficients[, c("Estimate", "Std. Error")]
}
mic <- model_impute(
  unclamped,
  model_fun = glm,
  rhs = "~ 0 + factor(year)",
  extractor = extractor,
  model_args = list(family = poisson)
)
mic$type <- "unclamped"
mic_clamped <- model_impute(
  clamped,
  model_fun = glm,
  rhs = "~ 0 + factor(year)",
  extractor = extractor,
  model_args = list(family = poisson)
)
mic_clamped$type <- "clamped"
mic <- rbind(mic, mic_clamped)
gsub(".*?([0-9]+)", "\\1", mic$Parameter) |>
  as.integer() -> mic$year
library(ggplot2)
ggplot(mic, aes(x = year, y = Estimate, ymin = LCL, ymax = UCL, color = type)) +
  geom_pointrange(position = position_dodge(width = 0.5)) +
  scale_x_continuous(breaks = seq_len(n_year)) +
  labs(
    x = "Year",
    y = "Estimated effect of year on log(count)",
    color = "Clamping"
  ) +
  theme_bw()
```
